Indexing Variation Graphs\thanks{Supported by the Wellcome Trust grant 098051.}

### 2016-10-13 paper/paper.tex
Yet another round of proofreading.

A \emph{string} $S[0, n-1] = s_{0} \dotsm s_{n-1}$ of length $\abs{S} = n$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{0, \dotsc, \sigma - 1}$. \emph{Text} strings $T[0, n-1]$ are terminated by an \emph{endmarker} $T[n-1] = \$ = 0$ not found anywhere else in the text. A \emph{substring} of string $S$ is a sequence of the form $S[i, j] = s_{i} \dotsm s_{j}$. We call substrings of the type $S[0, j]$ and $S[i, n-1]$ \emph{prefixes} and \emph{suffixes}, respectively, and substrings of length $k$ as \kmer{k}s. Substring $S[i, j]$ is a \emph{proper} substring of string $S$, if $S \ne S[i, j]$. We say that string $S'$ is a substring of string collection $\mathcal{S}$, if it is a substring of a string $S \in \mathcal{S}$.

@@

\begin{Definition}[de~Bruijn graph]
Let $k > 0$, and let $\mathcal{S}$ be a \kcollection{k}. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
\begin{itemize}
\item each node $v_{X} \in V$ represents a distinct \kmer{k} $X$ occurring in $\mathcal{S}$, with $G.\glabel(v_{X}) = X[0]$;
\item each node $v_{X} \in V$ has a \emph{key} $G.\gkey(v_{X}) = X$; and
\item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{k+1} $X[0]Y = XY[k-1]$ occurring in $\mathcal{S}$.
\end{itemize}
We use $\#^{k}$ as the source node $s$ and $\$^{k}$ as the sink node $t$, adding the edge $(t, s)$ in the usual way.
\end{Definition}

@@

The \emph{\kmer{k} index} is the simplest path index. It consists of a set of \emph{key-value pairs} $(X, V_{X})$, where $X \in \Sigma^{k}$ is a \kmer{k} and $V_{X} \subseteq V$ is the set of start nodes of the paths matching the \kmer{k}. If we store the pairs in a hash table, we can quickly search for patterns of length $k$. If we use binary search in a sorted list of pairs, queries become slower, but we gain the ability to search for patterns shorter than $k$ characters. The main drawback of these basic \kmer{k} indexes is their size, as they store the key-value pairs explicitly.

@@

Let $G' = (V', E')$ be a de~Bruijn graph of graph $G = (V, E)$, and let $\mathcal{S}$ be the \kcollection{k} used to define it. We attach a set of nodes of graph $G$ to each node $v' \in V'$ as a \emph{value} $G'.\gvalue(v)$:
$$
\set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.
$$
Apart from some technicalities near the source/sink nodes, $G'.\gvalue(v)$ is the set of start nodes of the paths matching pattern $G'.\gkey(v)$ in graph $G$. The index produces no \emph{false negatives} (path labels that exist in the input graph but not in the index). There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results in the input graph. See Figure~\ref{figure:graph-dbg} for an example.

@@

We can encode the de~Bruijn graph as a GCSA using $(\abs{V'} + \oh{\abs{V'}})(\log \sigma + 2)$ bits. By using a similar sampling scheme for the values as in the FM-index, we get a \kmer{k} index that uses a couple of bytes per \kmer{k} for a typical variation graphs (see Section~\ref{sect:experiments}). While this is much less than with the basic \kmer{k} indexes, it is still too much for large variation graphs.

@@

When a path of length $k' < k$ has a unique label, its extensions become \emph{redundant} nodes in an \orderk{k} de~Bruijn graph, if the path branches after the first $k'$ characters. By using shorter keys when possible, we can build a smaller graph equivalent to the de~Bruijn graph as a path index.

@@

\begin{Definition}[Maximally pruned graph] ~\\
Let $G'$ be a a pruned de~Bruijn graph of graph $G$ with key set $\mathcal{K}$. We say that $G'$ is \emph{maximally pruned}, if we cannot prune it any further using Lemma~\ref{lemma:dbg-prune}.
\end{Definition}

@@

We compared the query performance of the \orderk{128} GCSA2 to several FM-indexes for the reference sequence. SSA is the SDSL implementation (\texttt{csa\_wt<>} with SA sample period 17) of the \emph{succinct suffix array} \cite{Maekinen2005}, using a Huffman-shaped wavelet tree on top of the BWT. As the default FM-index in SDSL, it prioritizes query performance over compression. BWA is the FM-index in the \emph{Burrows-Wheeler Aligner} \cite{Li2009} (version~0.7.15 with SA sample period 32). Optimized for DNA sequences, BWA indexes both the reference and its reverse complement.

@@

GCSA2 uses denser SA sampling than the other indexes, with effective sample period 10.6. On the average, each sample contains 1.53~values and GCSA2 retrieves each distinct value 2.84~times, making the amount of work comparable to sample period 19.7. SSA with sample period 17 is 2.4x faster than GCSA2, highlighting the slowdown from checking multiple bitvectors $B_{c}$ in $\LF(i)$ queries, as well as from filtering out duplicates. BWA is closer to GCSA2 in $\locate$ performance, which can be explained by sparser sampling than in SSA. RLCSA is slower than the other indexes, but the difference is smaller than with $\find$ queries due to the optimizations for retrieving suffix array ranges. Assuming that $\locate$ queries in GCSA are 3x slower than RLCSA, as GCSA does use the optimizations, we can estimate that GCSA2 is 4x faster than GCSA.

@@

GCSA2 is a path index for variation graphs. It uses a de~Bruijn graph as a \kmer{k} index of the variation graph, prunes it by merging redundant subgraphs, and encodes the result with a generalization of the FM\nobreakdash-index. The index supports queries of length up to $k$ exactly and longer queries with false positives. GCSA2 also includes extensions based on suffix trees; other extensions have been considered but not implemented (see Appendix~\ref{appendix:hypertext}). The index is used in the variation graph toolkit vg for e.g.~read alignment based on maximal exact matches.

@@

\begin{proof}[Lemma~\ref{lemma:pg-context}: Context length]
If $\abs{X} \le 1$, the statement is true by definition for $m = \abs{X}$. Now let $M_{i+1} = G'.\find(X[i+1, \abs{X}-1])$ be the set of all nodes $v' \in V'$ such that substring $X[i+1, i+m]$ is a prefix of key $G'.\gkey(v')$, and assume that the set is nonempty.

Consider the set
$$
M_{i} = \bigcup_{v' \in M_{i+1}} G'.\gpred(v', X[i]) = G'.\find(X[i, \abs{X}-1]).
$$
There is an edge $(u', v') \in E'$ if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$. Hence key $G'.\gkey(u')$ prefix-matches string $X[i, i+m]$ for all nodes $u' \in M_{i}$.

Now let $u' \in V'$ be a node with key $G'.\gkey(u')$ prefix-matching string $X[i, i+m]$. If the key is a prefix of string $X[i, i+m]$, there is an edge $(u', v') \in E'$ to all nodes $v' \in M_{i+1}$, and hence $u' \in M_{i}$. Otherwise let $S[j, j+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring of the \kcollection{k} used for building the path graph. As $S[j+1, j+m] = X[i+1, i+m]$, the substring starting at $S[j+1]$ is represented by a node $v' \in M_{i+1}$, and hence $u' \in M_{i}$.

Set $M_{i}$ is the set of all nodes $u' \in V'$ such that substring $X[i, i+m]$ prefix-matches key $G'.\gkey(u')$. If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$. Otherwise the context length becomes $\min(m+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
\end{proof}

@@

Let $P'' = v''_{0} \dotsm v''_{\abs{P''}-1}$ be a in graph $G''$. We transform it into an equivalent path $P' = v'_{0} \dotsm v'_{\abs{P''}-1}$ in graph $G'$. There are two cases for $\abs{P''} = 1$. If $v''_{0} = v''_{K}$, we can replace it with any $v'_{0} \in V'_{K}$. Otherwise we use the node $v'_{0} \in V'$ such that $G'.\gkey(v'_{0}) = G''.\gkey(v''_{0})$.

@@

\item If key $G'.\gkey(v')$ is a proper prefix of string $Y$, there is only one possible predecessor $v' = v'_{0}$. Because key $G''.\gkey(v''_{0})$ prefix-matches string $Y$, it must also be a proper prefix of the string. Hence either $G'.\gkey(v'_{0}) = G''.\gkey(v''_{0})$ or $v'_{0} \in V'_{K}$ and $v''_{0} = v''_{K}$.
\end{enumerate}
The shorter of $Y$ and $G''.\gkey(v''_{0})$ can be used as $K_{0}$.

@@

Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$ for an edge $(u', v') \in E'$. Then string $G'.\glabel(u') \cdot G'.\gkey(v')$ must then be a prefix of key $G'.\gkey(u')$. There cannot be edges $(u', w')$ to other nodes $w' \ne v'$, as keys $G'.\gkey(v')$ and $G'.\gkey(w')$ would prefix-match.

Let $S[i, i+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring in the \kcollection{k}. Because graph $G'$ is a pruned de~Bruijn graph and because of the assumption on key lengths, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in the \kcollection{k} is $G'.\gvalue(v')$. As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ over all nodes $v \in G'.\gvalue(v')$.

@@

Query $G'.\locate(X)$ retrieves the values $G'.\gvalue(v')$ for nodes $v' \in G'.\find(X)$ and filters out duplicates. Instead of storing the values explicitly for all nodes, GCSA uses a \emph{sampling} scheme to save space. We assume that the nodes $v \in V$ of the input graph $G = (V, E)$ are conveniently chosen integers. If $(u, v) \in E$ is the only outgoing edge from node $u$ and the only incoming edge to node $v$, it should be that $v = u+1$.

@@

Let $B_{S}[0, \abs{V'}-1]$ be a bitvector. If we have sampled the values for the node $v' \in V'$ with lexicographic rank $i$, we mark that as $B_{S}[i] = 1$. We can then determine the rank of node $v'$ among the sampled nodes as $j = B_{S}.\rank(i, 1)$. For each sampled node $v' \in V$, we store the size of the value set $\abs{G'.\gvalue(v')}$ in another bitvector $B_{V}$, using the same encoding as in bitvectors $\bvIN$ and $\bvOUT$. We store the samples in array $V_{S}$ in the same order, using $\log \abs{V}$ bits each. The sampled values for node $v'$ with rank $j$ among the sampled nodes can be found at $V_{S}[B_{V}.\select(j, 1) + 1, B_{V}.\select(j+1, 1)]$.

@@

Because we store path labels explicitly, we only do a limited number of \emph{doubling steps}, typically $2$ or $3$. After $d$ doubling steps, the length of the paths is $2^{d} k$, and we can use them to build a maximally pruned \orderk{(2^{d} k)} de~Bruijn graph. Each doubling step consists of a \emph{pruning} step, followed by an \emph{extension} step. The pruning step applies a limited form of Lemma~\ref{lemma:dbg-prune} to lexicographic ranges of paths. Given two paths $P$ and $P'$, we can determine the longest common prefix of the path labels by using the keys $P.\gkey$ and $P'.\gkey$ and the LCP array. If all paths sharing a prefix start from the same node, we merge them into a single path $Q$ with $Q.\gkey$ based on the shared prefix and $Q.\gext = -1$.

@@

The doubling steps are followed by the \emph{merging step}, which transforms the paths into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$. We merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$. If paths $P_{0}, \dotsm, P_{m-1}$ all have the same key, we create a node $v' \in V'$ with the shared key as $G'.\gkey(v')$ and with $G'.\gvalue(v') = \bigcup_{i=0}^{m-1} P_{i}.\gvalue$. We also store the union of predecessor labels as $G'.\gpred(v') = \bigcup_{i=0}^{m-1} P_{i}.\gpred$. We then apply Lemma~\ref{lemma:dbg-prune} maximally, transforming graph $G'$ into graph $G''$.

Storing the paths and graphs may require hundreds of gigabytes of memory when indexing whole-genome variation graphs. To avoid that, we keep them on \emph{disk} when possible. The subgraph corresponding to each chromosome is stored in a separate file, with paths sorted by their labels in lexicographic order. Extension steps are done separately for each of the chromosomes. A pruning step merges the sorted files. It keeps reading paths into a buffer, until it has found a maximal range of paths that can be merged. The merged path is written into the new file for that chromosome, and the original range of paths is removed from the buffer. The merging step works in a similar way.

After creating the nodes $V''$ of the maximally pruned \orderk{k'} de~Bruijn graph $G'' = (V'', E'')$, we \emph{build the index}. Sequences $\BWT$ and $\bvIN$ can be generated from the predecessor sets $G''.\gpred(v'')$, while the outdegree sequence $\bvOUT$ requires further processing. There is an edge $(u'', v'') \in E''$ if and only if string $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ and $c \in G''.\gpred(v')$. We determine the edges and produce the outdegree sequence by scanning the node file sequentially with $\sigma+1$ pointers. The pointer corresponding to node $v''$ scans the entire file, while each of the remaining $\sigma$ pointers scans only the range of nodes $u''$ with $G''.\glabel(u'') = c$ for a character $c \in \Sigma$. We also sample the nodes for $\locate$ queries during the same scans.

@@

Let $G' = (V', E')$ be a path graph with $G'.\gpred(v', c) \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$. This is true for de~Bruijn graphs, and also for maximally pruned de~Bruijn graphs (Lemmas~\ref{lemma:dbg-maximal} and \ref{lemma:pg-keys}). We can use the simplified encoding of the original GCSA \cite{Siren2014} for the GCSA of such path graphs.

We replace sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$. If node $v' \in V$ with lexicographic rank $i$ has a predecessor with label $c \in \Sigma$, we set $B_{c}[i] = 1$. The backward step becomes:
\begin{align*}
sp_{out} & = C[X[i]] + B_{X[i]}.\rank(sp_{i+1}, 1); \\
ep_{out} & = C[X[i]] + B_{X[i]}.\rank(ep_{i+1}+1, 1) - 1; \\
[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
\end{align*}
Two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) are replaced with a cheap $B_{c}.\rank()$.

@@

Let $G' = (V', E')$ be an \orderk{k} path graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order. The \emph{LCP array} of graph $G'$ is an array $\LCP[0, \abs{V'}-1]$, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$. If we build a trie of keys $G'.\gkey(v')$ for $v' \in V'$ and compact the unary paths into single edges, the resulting tree can be considered the \orderk{k} \emph{suffix tree} of graph $G'$.

@@

If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x \log_{x} \abs{V'}}$ time with $\Oh{\log_{x} \abs{V'}}$ random memory accesses, while using $\frac{x}{x-1} \abs{V'} \log k$ bits of space. Because the nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use. By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern. Hence we can safely use $\parent$ queries with such ranges.\footnote{The $\textsf{shorter}$ query in the variable-order de~Bruijn graph \cite{Boucher2014} is essentially a $\parent$ query.}

@@

For value counting, we use array $R$ to store the number of redundant values as above. We use another array $A[0, \abs{V'}-1]$ to store the number of additional values in each node $v'_{i} \in V'$ as $A[i] = \abs{G'.\gvalue(v'_{i})}-1$, and encode it as a bitvector $B_{A}$ in the same way as array $R$ above. The number of distinct values in range $[sp,ep] = G'.\find(X)$ for a pattern $X$ is
\begin{align*}
G'.\countq(X) & = G'.\countq([sp, ep]) \\
 & = \sum_{i = sp}^{ep} (A[i] + 1) - \sum_{i = sp}^{ep - 1} R[i].
\end{align*}
The bitvectors are often highly compressible \cite{Gagie2015}, but GCSA already uses one of the compression schemes implicitly when it prunes the de~Bruijn graph.

>>>>

A \emph{string} $S[0, n-1] = s_{0} \dotsm s_{n-1}$ of length $\abs{S} = n$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{0, \dotsc, \sigma - 1}$. \emph{Text} strings $T[0, n-1]$ are terminated by an \emph{endmarker} $T[n-1] = \$ = 0$ not found anywhere else in the text. A \emph{substring} of string $S$ is a sequence of the form $S[i, j] = s_{i} \dotsm s_{j}$. We call substrings of the type $S[0, j]$ and $S[i, n-1]$ \emph{prefixes} and \emph{suffixes}, respectively, and refer to substrings of length $k$ as \kmer{k}s. Substring $S[i, j]$ is a \emph{proper} substring of string $S$, if $S \ne S[i, j]$. We say that string $S'$ is a substring of string collection $\mathcal{S}$, if it is a substring of a string $S \in \mathcal{S}$.

@@

\begin{Definition}[de~Bruijn graph]
Let $k > 0$, and let $\mathcal{S}$ be a \kcollection{k}. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
\begin{itemize}
\item each node $v_{X} \in V$ represents a distinct \kmer{k} $X$ occurring in $\mathcal{S}$, with $G.\glabel(v_{X}) = X[0]$;
\item each node $v_{X} \in V$ has a \emph{key} $G.\gkey(v_{X}) = X$; and
\item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{(k+1)} $X[0]Y = XY[k-1]$ occurring in $\mathcal{S}$.
\end{itemize}
We use $\#^{k}$ as the source node $s$ and $\$^{k}$ as the sink node $t$, adding the edge $(t, s)$ in the usual way.
\end{Definition}

@@

The \emph{\kmer{k} index} is the simplest path index. It consists of a set of \emph{key-value pairs} $(X, V_{X})$, where $X \in \Sigma^{k}$ is a \kmer{k} and $V_{X} \subseteq V$ is the set of the start nodes of the paths matching the \kmer{k}. If we store the pairs in a hash table, we can quickly search for patterns of length $k$. If we use binary search in a sorted list of pairs, queries become slower, but we gain the ability to search for patterns shorter than $k$ characters. The main drawback of these basic \kmer{k} indexes is their size, as they store the key-value pairs explicitly.

@@

Let $G' = (V', E')$ be a de~Bruijn graph of graph $G = (V, E)$, and let $\mathcal{S}$ be the \kcollection{k} used to define it. We attach a set of nodes of graph $G$ to each node $v' \in V'$ as a \emph{value} $G'.\gvalue(v)$:
$$
\set{ \mathcal{S}.\gnode(S, i) \mid S \in \mathcal{S}, S[i, i+k-1] = G'.\gkey(v)}.
$$
Apart from some technicalities near the source/sink nodes, $G'.\gvalue(v)$ is the set of the start nodes of the paths matching pattern $G'.\gkey(v)$ in graph $G$. The index produces no \emph{false negatives} (path labels that exist in the input graph but not in the index). There may be \emph{false positives} (path labels that exist in the index but not in the input graph) with patterns longer than $k$, but we can avoid them by \emph{verifying} the results in the input graph. See Figure~\ref{figure:graph-dbg} for an example.

@@

We can encode the de~Bruijn graph as a GCSA using $(\abs{V'} + \oh{\abs{V'}})(\log \sigma + 2)$ bits. By using a similar sampling scheme for the values as in the FM-index, we get a \kmer{k} index that uses a couple of bytes per \kmer{k} for typical variation graphs (see Section~\ref{sect:experiments}). While this is much less than with the basic \kmer{k} indexes, it is still too much for large variation graphs.

@@

When a path of length $k' < k$ has a unique label, its extensions become \emph{redundant} nodes in the \orderk{k} de~Bruijn graph, if the path branches after the first $k'$ characters. By using shorter keys when possible, we can build a smaller graph that is equivalent to the de~Bruijn graph as a path index.

@@

\begin{Definition}[Maximally pruned graph] ~\\
Let $G'$ be a a pruned de~Bruijn graph of graph $G$. We say that $G'$ is \emph{maximally pruned}, if we cannot prune it any further using Lemma~\ref{lemma:dbg-prune}.
\end{Definition}

@@

We compared the query performance of the \orderk{128} GCSA2 to several FM-indexes for the reference sequence. SSA is the SDSL implementation (\texttt{csa\_wt<>} with SA sample period 17) of the \emph{succinct suffix array} \cite{Maekinen2005}, using a Huffman-shaped wavelet tree on top of the BWT. As the default FM-index in SDSL, it prioritizes query performance over compression. BWA is the FM-index in the \emph{Burrows-Wheeler Aligner} \cite{Li2009} (version~0.7.15 with the default SA sample period 32). Optimized for DNA sequences, BWA indexes both the reference and its reverse complement.

@@

GCSA2 uses denser SA sampling than the other indexes, with effective sample period 10.6. On the average, each sample contains 1.53~values, and GCSA2 retrieves each distinct value 2.84~times, making the amount of work comparable to sample period 19.7. SSA with sample period 17 is 2.4x faster than GCSA2, highlighting the slowdown from checking multiple bitvectors $B_{c}$ in $\LF(i)$ queries, as well as from filtering out duplicates. BWA is closer to GCSA2 in $\locate$ performance, which can be explained by sparser sampling than in SSA. RLCSA is slower than the other indexes, but the difference is smaller than with $\find$ queries due to the optimizations for retrieving suffix array ranges. Assuming that $\locate$ queries in GCSA are 3x slower than RLCSA, as GCSA does use the optimizations, we can estimate that GCSA2 is 4x faster than GCSA.

@@

GCSA2 is a path index for variation graphs. It uses a de~Bruijn graph as a \kmer{k} index of the variation graph, prunes it by merging redundant subgraphs, and encodes the result with a generalization of the FM\nobreakdash-index. The index supports queries of length up to $k$ exactly, and longer queries with false positives. GCSA2 also includes extensions based on suffix trees; other extensions have been considered but not implemented (see Appendix~\ref{appendix:hypertext}). The index is used in the variation graph toolkit vg for e.g.~read alignment based on maximal exact matches.

@@

\begin{proof}[Lemma~\ref{lemma:pg-context}: Context length]
If $\abs{X} \le 1$, the statement is true by definition for $m = \abs{X}$. Now let $M_{i+1} = G'.\find(X[i+1, \abs{X}-1])$ be the set of all nodes $v' \in V'$ such that substring $X[i+1, i+m_{i+1}]$ is a prefix of key $G'.\gkey(v')$, and assume that the set is nonempty.

Consider the set
$$
M_{i} = \bigcup_{v' \in M_{i+1}} G'.\gpred(v', X[i]) = G'.\find(X[i, \abs{X}-1]).
$$
There is an edge $(u', v') \in E'$ if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$. Hence key $G'.\gkey(u')$ prefix-matches string $X[i, i+m_{i+1}]$ for all nodes $u' \in M_{i}$.

Now let $u' \in V'$ be a node with key $G'.\gkey(u')$ prefix-matching string $X[i, i+m_{i+1}]$. If the key is a prefix of string $X[i, i+m_{i+1}]$, there is an edge $(u', v') \in E'$ to all nodes $v' \in M_{i+1}$, and hence $u' \in M_{i}$. Otherwise let $S[j, j+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring of the \kcollection{k} used for building the path graph. As $S[j+1, j+m_{i+1}] = X[i+1, i+m_{i+1}]$, the substring starting at $S[j+1]$ is represented by a node $v' \in M_{i+1}$, and hence $u' \in M_{i}$.

Set $M_{i}$ is the set of all nodes $u' \in V'$ such that substring $X[i, i+m_{i+1}]$ prefix-matches key $G'.\gkey(u')$. If $\abs{M_{i}} > 1$, string $X[i, i+m_{i+1}]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can set $m_{i} = m_{i+1}+1$. Otherwise we set $m_{i} = \min(m_{i+1}+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
\end{proof}

@@

Let $P'' = v''_{0} \dotsm v''_{\abs{P''}-1}$ be a in graph $G''$. We transform it into an equivalent path $P' = v'_{0} \dotsm v'_{\abs{P''}-1}$ in graph $G'$. There are two cases for $\abs{P''} = 1$. If $v''_{0} = v''_{K}$, we can replace it with any $v'_{0} \in V'_{K}$. Otherwise we use the node $v'_{0} \in V'$ with $G'.\gkey(v'_{0}) = G''.\gkey(v''_{0})$.

@@

\item If key $G'.\gkey(v')$ is a proper prefix of string $Y$, there is only one possible predecessor $v' = v'_{0}$. Because key $G''.\gkey(v''_{0})$ prefix-matches string $Y$, it must also be a proper prefix of the string. Hence either $G'.\gkey(v'_{0}) = G''.\gkey(v''_{0})$ or $v'_{0} \in V'_{K}$ and $v''_{0} = v''_{K}$.
\end{enumerate}
In every case, $G'.\gvalue(v'_{0}) = G''.\gvalue(v''_{0})$, and we can use shorter of strings $Y$ and $G''.\gkey(v''_{0})$ as $K_{0}$.

@@

Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$ for an edge $(u', v') \in E'$. String $G'.\glabel(u') \cdot G'.\gkey(v')$ must then be a prefix of key $G'.\gkey(u')$. There cannot be edges $(u', w')$ to other nodes $w' \ne v'$, as keys $G'.\gkey(v')$ and $G'.\gkey(w')$ would prefix-match.

Let $S[i, i+\abs{G'.\gkey(u')}-1] = G'.\gkey(u')$ be a substring of $\mathcal{S}$. Because $G'.\gkey(v')$ is a substring of $G'.\gkey(u')$ and $G'$ is a pruned de~Bruijn graph, the set of nodes $\mathcal{S}.\gnode(S, i+1)$ over all occurrences of substring $G'.\gkey(u')$ in $\mathcal{S}$ is $G'.\gvalue(v')$. As node $u'$ has no other successors, set $G'.\gvalue(u')$ is the union of sets $G.\gpred(v, G'.\glabel(u'))$ over all nodes $v \in G'.\gvalue(v')$.

@@

Query $G'.\locate(X)$ retrieves the values $G'.\gvalue(v')$ for nodes $v' \in G'.\find(X)$ and filters out duplicates. Instead of storing the values explicitly for all nodes, GCSA uses a \emph{sampling} scheme to save space. We assume that the nodes of the input graph $G = (V, E)$ are conveniently chosen integers. If $(u, v) \in E$ is the only outgoing edge from node $u$ and the only incoming edge to node $v$, it should be that $v = u+1$.

@@

Let $B_{S}[0, \abs{V'}-1]$ be a bitvector. If we have sampled the values for the node $v' \in V'$ with lexicographic rank $i$, we mark that as $B_{S}[i] = 1$. We can then determine the rank of node $v'$ among the sampled nodes as $j = B_{S}.\rank(i, 1)$. For each sampled node $v' \in V$, we store the size of the value set $\abs{G'.\gvalue(v')}$ in another bitvector $B_{V}$, using the same encoding as for bitvectors $\bvIN$ and $\bvOUT$. We store the samples in array $V_{S}$ in the same order, using $\log \abs{V}$ bits each. The sampled values for node $v'$ with rank $j$ among the sampled nodes can be found at $V_{S}[B_{V}.\select(j, 1) + 1, B_{V}.\select(j+1, 1)]$.

@@

Because we store path labels explicitly, we only do a limited number of \emph{doubling steps}, typically $2$ or $3$. After $d$ doubling steps, the length of the paths is $2^{d} k$, and we can use them to build a maximally pruned \orderk{(2^{d} k)} de~Bruijn graph. Each doubling step consists of a \emph{pruning} step, followed by an \emph{extension} step. The pruning step applies a limited form of Lemma~\ref{lemma:dbg-prune} to lexicographic ranges of paths. Given two paths $P$ and $P'$, we can determine the length of the longest common prefix of the path labels by using the keys $P.\gkey$ and $P'.\gkey$ and the LCP array. If all paths sharing a prefix start from the same node, we merge them into a single path $Q$ with $Q.\gkey$ based on the shared prefix and $Q.\gext = -1$.

@@

The doubling steps are followed by the \emph{merging step}, which transforms the paths into the nodes of a maximally pruned de~Bruijn graph $G'' = (V'', E'')$. We merge the paths with identical keys into the nodes of a pruned de~Bruijn graph $G' = (V', E')$. If paths $P_{0}, \dotsm, P_{m-1}$ all have the same key, we create a node $v' \in V'$ with the shared key as $G'.\gkey(v')$ and with $G'.\gvalue(v') = \bigcup_{i=0}^{m-1} P_{i}.\gvalue$. We also store the union of predecessor labels as $v'.\gpred = \bigcup_{i=0}^{m-1} P_{i}.\gpred$. We then apply Lemma~\ref{lemma:dbg-prune} maximally, transforming graph $G'$ into graph $G''$.

Storing the paths and the graphs may require hundreds of gigabytes of memory when indexing whole-genome variation graphs. To avoid that, we keep them on \emph{disk} when possible. The subgraph corresponding to each chromosome is stored in a separate file, with paths sorted by their labels in lexicographic order. Extension steps are done separately for each of the chromosomes. The pruning step merges the sorted files. It keeps reading paths into a buffer, until it has found a maximal range of paths that can be merged. The merged path is written into the new file for that chromosome, and the original range of paths is removed from the buffer. The merging step works in a similar way.

After creating the nodes $V''$ of the maximally pruned \orderk{k'} de~Bruijn graph $G'' = (V'', E'')$, we \emph{build the index}. Sequences $\BWT$ and $\bvIN$ can be generated from the predecessor sets $v''.\gpred$, while the outdegree sequence $\bvOUT$ requires further processing. There is an edge $(u'', v'') \in E''$ if and only if string $c \cdot G''.\gkey(v'')$ prefix-matches key $G''.\gkey(u'')$ and $c \in v''.\gpred$. We determine the edges and produce the outdegree sequence by scanning the node file sequentially with $\sigma+1$ pointers. The pointer corresponding to node $v''$ scans the entire file, while each of the remaining $\sigma$ pointers scans only the range of nodes $u''$ with $G''.\glabel(u'') = c$ for a character $c \in \Sigma$. We also sample the nodes for $\locate$ queries during the same scans.

@@

Let $G' = (V', E')$ be a path graph with $G'.\gpred(v', c) \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$. This is true for de~Bruijn graphs, and also for maximally pruned de~Bruijn graphs (Lemmas~\ref{lemma:dbg-maximal} and \ref{lemma:pg-keys}). We can use the simplified encoding of the original GCSA \cite{Siren2014} with such path graphs.

We replace the sequences $\BWT$ and $\bvIN$ with \emph{indicator bitvectors} $B_{c}[0, \abs{V'}-1]$ for all $c \in \Sigma$. If node $v' \in V$ with lexicographic rank $i$ has a predecessor with label $c \in \Sigma$, we set $B_{c}[i] = 1$. The backward step becomes:
\begin{align*}
sp_{out} & = C[X[i]] + B_{X[i]}.\rank(sp_{i+1}, 1); \\
ep_{out} & = C[X[i]] + B_{X[i]}.\rank(ep_{i+1}+1, 1) - 1; \\
[sp_{i}, ep_{i}] & = [\bvOUT.\rank(sp_{out}, 1), \bvOUT.\rank(ep_{out}, 1)].
\end{align*}
Two expensive queries ($\bvIN.\select()$ and $\BWT.\rank()$) are replaced with a cheap $B_{c}.\rank()$.

@@

Let $G' = (V', E')$ be an \orderk{k} path graph, and let $v'_{0}, \dotsc, v'_{\abs{V'}-1}$ be its nodes in lexicographic order. The \emph{LCP array} of graph $G'$ is an array $\LCP[0, \abs{V'}-1]$, where $\LCP[i]$ is the length of the longest common prefix of keys $G'.\gkey(v'_{i-1})$ and $G'.\gkey(v'_{i})$, with $\LCP[0] = 0$. If we build a trie of keys $G'.\gkey(v')$ for $v' \in V'$ and compact the unary paths into single edges, the resulting tree can be considered the \emph{suffix tree} of graph $G'$.

@@

If we build an $x$\nobreakdash-ary tree over the LCP array, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x \log_{x} \abs{V'}}$ time with $\Oh{\log_{x} \abs{V'}}$ random memory accesses, while using $\frac{x}{x-1} \abs{V'} \log k$ bits of space. Because the nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use. By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern. Hence we can safely use $\parent$ queries with such ranges.\footnote{The $\textsf{shorter}$ query in the variable-order de~Bruijn graph \cite{Boucher2014} is essentially a $\parent$ query.}

@@

For value counting, we use array $R$ to store the number of redundant values as above. We use another array $A[0, \abs{V'}-1]$ to store the number of additional values in each node $v'_{i} \in V'$ as $A[i] = \abs{G'.\gvalue(v'_{i})}-1$, and encode it as a bitvector $B_{A}$ in the same way as array $R$ above. The number of distinct values in range $[sp,ep] = G'.\find(X)$ for a pattern $X$ is
\begin{align*}
G'.\countq(X) & = G'.\countq(sp, ep) \\
 & = \sum_{i = sp}^{ep} (A[i] + 1) - \sum_{i = sp}^{ep - 1} R[i].
\end{align*}
The bitvectors are often highly compressible \cite{Gagie2015}, but GCSA already uses one of the compression schemes implicitly when it prunes the de~Bruijn graph.
